Quantifying the rock-paper-scissors effect

### 2023-10-31 inst/plackett_luce_monster.Rmd
label all chunks, cache the slow ones
consistentification

---
title: "A generalization of Plackett-Luce likelihoods to incorporate non-independence of competitors"
output: html_document
date: '2022-05-16'
---

@@

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

Here I modify standard Plackett-Luce likelihood functions to include
an element of non-independence among the competitors.  There seem to
be a few distinct statistical models for nonindependence:

@@

```{r}
f(4.4,3.3, 0.2)*12
f(4.4,3.3, 0.0)*12
```

```{tikz, fig.cap = "Partial probability tree for five competitors $a-e$ with $a$ supporting $b$, hyper3 approach", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2,3); 
\draw (0,0) -- (2,2);
\draw (0,0) -- (2,1);
\draw (0,0) -- (2,0);
\draw (0,0) -- (2,-1);
\node at (0,2.5) (eq1) {$a\longrightarrow b$};

@@

```{tikz, fig.cap = "Partial probability tree for five competitors $a-e$ with mutually supporting  group $abc$, hyper3 approach.  If any competitor in the set {$a,b,c$} finishes, they lend their strength to the other competitors in the set who are still running", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2,3); 
\draw (0,0) -- (2,2);
\draw (0,0) -- (2,1);
\draw (0,0) -- (2,0);
\draw (0,0) -- (2,-1);
\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}$};

@@

```{tikz, fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $\\left\\lbrace a,b,c\\right\\rbrace$ [with support term $\\lambda$] and $(de)$ [with support term $\\mu$], hyper3 approach", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2,3); 
\draw (0,0) -- (2,2);
\draw (0,0) -- (2,1);
\draw (0,0) -- (2,0);
\draw (0,0) -- (2,-1);
\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}\underbrace{\left\lbrace d,e\right\rbrace}_{\mu}$};

@@

```{r}
a <- read.table("timesData.csv",sep=",", header=TRUE)
wanted <- c("California Institute of Technology", "Harvard University", 
"Massachusetts Institute of Technology", "Princeton University", 
"Stanford University", "University of Cambridge", "University of Oxford")
names(wanted) <- c("cal","harv","mass","prin","stan","cam","ox")

>>>>

---
title: "A generalization of Plackett-Luce likelihoods to incorporate non-independence of competitors"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: jss_hyper3.bib
---

@@

```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

To cite the `hyper2` package in publications, please use @hankin2017a.
Here I modify standard Plackett-Luce likelihood functions to include
an element of non-independence among the competitors.  There seem to
be a few distinct statistical models for nonindependence:

@@

```{r evalat44}
f(4.4,3.3, 0.2)*12
f(4.4,3.3, 0.0)*12
```

```{tikz, label=tikzpartprobtree1,fig.cap = "Partial probability tree for five competitors $a-e$ with $a$ supporting $b$, hyper3 approach", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2,3); 
\draw (0,0) -- (2,2);
\draw (0,0) -- (2,1);
\draw (0,0) -- (2,0);
\draw (0,0) -- (2,-1);
\node at (0,2.5) (eq1) {$a\longrightarrow b$};

@@

```{tikz, label=tikzpic2, fig.cap = "Partial probability tree for five competitors $a-e$ with mutually supporting  group $abc$, hyper3 approach.  If any competitor in the set {$a,b,c$} finishes, they lend their strength to the other competitors in the set who are still running", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2,3); 
\draw (0,0) -- (2,2);
\draw (0,0) -- (2,1);
\draw (0,0) -- (2,0);
\draw (0,0) -- (2,-1);
\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}$};

@@

```{tikz, label=tikz3,fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $\\left\\lbrace a,b,c\\right\\rbrace$ [with support term $\\lambda$] and $(de)$ [with support term $\\mu$], hyper3 approach", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2,3); 
\draw (0,0) -- (2,2);
\draw (0,0) -- (2,1);
\draw (0,0) -- (2,0);
\draw (0,0) -- (2,-1);
\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}\underbrace{\left\lbrace d,e\right\rbrace}_{\mu}$};

@@

```{r, label=readtableandcalc,cache=TRUE}
a <- read.table("timesData.csv",sep=",", header=TRUE)
wanted <- c("California Institute of Technology", "Harvard University", 
"Massachusetts Institute of Technology", "Princeton University", 
"Stanford University", "University of Cambridge", "University of Oxford")
names(wanted) <- c("cal","harv","mass","prin","stan","cam","ox")

### 2023-10-31 inst/podium.Rmd
consistentification

---
title: "Does being on the podium affect Formula 1 drivers' performance in the next race?"
output: bookdown::html_document2
---

@@

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

This short document defines the `podium()` function which tests the
hypothesis that a Formula 1 driver's being on the podium at one venue
affects his performance on the next.

>>>>

---
title: "Does being on the podium affect Formula 1 drivers' performance in the next race?"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: jss_hyper3.bib
---

@@

```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

To cite the `hyper2` package in publications, please use @hankin2017a.
This short document defines the `podium()` function which tests the
hypothesis that a Formula 1 driver's being on the podium at one venue
affects his performance on the next.  Data from the 2007 season are
used.  It uses `hyper3` objects from the `hyper2` package.  A detailed
use-case for the top six drivers (Alonso, Hamilton, Heidfeld, Kubica,
Massa, and Raikkonen), and only the first five venues (Austria,
Malaysia, Bahrain, Spain, and Monaco) is given.

***TLDR***: Statistically significant evidence for the podium effect
was found, with about 2.7 units of support for the alternative.

Function `podium()`, defined below, takes a race table and a value of
$\lambda$ and returns a `hyper3` support function.

### 2023-10-31 inst/powerboat.Rmd
consistentification

---
title: "powerboat"
output: html_document
---

@@

[powerboat](https://en.wikipedia.org/w/index.php?title=2018_F1_Powerboat_World_Championship&oldid=920386507)
racing is an important and prestigious sport.  Here I analyse a
dataset taken from the 2018 F1 Powerboat world championships.

>>>>

---
title: "powerboats with `hyper2`"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: jss_hyper3.bib
---

---

@@

```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

To cite the `hyper2` package in publications, please use @hankin2017a.
[powerboat](https://en.wikipedia.org/w/index.php?title=2018_F1_Powerboat_World_Championship&oldid=920386507)
racing is an important and prestigious sport.  Here I analyse a
dataset taken from the 2018 F1 Powerboat world championships.

### 2023-10-31 inst/race3.Rmd
do not need to load disordR
consistentification

---
title: "Races with hyper3 objects"
author: "Robin KS Hankin"
date: "10/04/2022"
output: html_document
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("disordR")
```

(takes about an hour to process without cache)

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

Consider a race in which there are six runners 1-6 but we happen to
know that three of the runners (1,2,3) are clones of strength $p_a$,
say, two of the runners (4,5) have strength $p_b$, and the final
runner (6) is of strength $p_c$.  The runners race and the finishing order is:

>>>>

---
title: "Races with hyper3 objects"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: jss_hyper3.bib
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

(takes about an hour to process without cache).  To cite the `hyper2`
package in publications, please use @hankin2017a.  Consider a race in
which there are six runners 1-6 but we happen to know that three of
the runners (1,2,3) are clones of strength $p_a$, say, two of the
runners (4,5) have strength $p_b$, and the final runner (6) is of
strength $p_c$.  The runners race and the finishing order is:

### 2023-10-31 inst/retain.Rmd
consistentification

---
title: "Keeping and discarding players"
output: html_document
---

@@

## Functions keep() and keep_flawed()

Function `keep_flawed()` can be confusing.  Originally, the package
included a function `keep()` but this used flawed logic, hence the
name of its replacement.  Here is a tiny example that should clarify
things.  We will use a subset of `skating_table`:

>>>>

---
title: "Keeping and discarding players: functions `keep()` and `keep_flawed()`"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: jss_hyper3.bib
---

@@

```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

To cite the `hyper2` package in publications, please use @hankin2017a.
Function `keep_flawed()` can be confusing.  Originally, the package
included a function `keep()` but this used flawed logic, hence the
name of its replacement.  Here is a tiny example that should clarify
things.  We will use a subset of `skating_table`:

### 2023-10-31 inst/rock_paper_scissors_monster.Rmd
label all chunks, cache the slow ones
consistentification

---
title: "Quantifying the rock-paper-scissors effect"
output: html_document
---

@@

(takes about an hour to run without cache)

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

## Abstract

Here I use the `hyper2` package (Hankin 2017) to quantify the strength
of a circular rock-paper-scissors effect using the `hyper2` package,
improving on previous results (West and Hankin 2008) which merely
established its existence.  I use two datasets, a three-player
scoreline and a synthetic dataset referring to wear on a seven-tooth
gear wheel.

@@

The existence of a rock-paper-scissors relationship between the
players was discussed in Hankin (2010) and the null of no RPS
phenomena could be rejected using the Alymer test.

@@

```{r}
maxp(H)
```

@@

## References

* L. J. West and R. K. S. Hankin 2008.  "Exact tests for two-way
  contingency tables with structural zeros".  _Journal of Statistical
  Software_ 2008, volume 28 issue 11.
* R. K. S. Hankin 2010.  "A generalization of the Dirichlet
  distribution".  _Journal of Statistical Software_ 2010, volume 33
  issue 11.
* R. K. S. Hankin 2017.  "Partial rank data with the `hyper2` package:
  likelihood functions for generalized Bradley-Terry models".  _The R
  Journal_, volume 9, number 2.

### Package dataset

>>>>

---
title: "Quantifying the rock-paper-scissors effect"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: jss_hyper3.bib
---

@@

```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

To cite the `hyper2` package in publications, please use @hankin2017a;
takes about an hour to run without cache.  Here I use the `hyper2`
package to quantify the strength of a circular rock-paper-scissors
effect using the `hyper2` package, improving on previous results
[@west2008] which merely established its existence.  I use two
datasets, a three-player scoreline and a synthetic dataset referring
to wear on a seven-tooth gear wheel.

@@

The existence of a rock-paper-scissors relationship between the
players was discussed in @hankin2010 and the null of no RPS phenomena
could be rejected using the Alymer test.

@@

```{r,calcmaxp,cache=TRUE}
maxp(H)
```

@@

### Package dataset {-}
